---
title: "Repertoire Analysis"
author: "Nick Borcherding"
date: "8/21/2020"
output: html_document
---
# Loading Data

## Read filtered contig
```{r}
S1.contig <- read.csv("~/Documents/GitHub/HNSCC_Descriptive/Data/VDJ_data/Sample1_filtered_contig_annotations.csv")
S2.contig <- read.csv("~/Documents/GitHub/HNSCC_Descriptive/Data/VDJ_data/Sample2_filtered_contig_annotations.csv")
S3.contig <- read.csv("~/Documents/GitHub/HNSCC_Descriptive/Data/VDJ_data/Sample3_filtered_contig_annotations.csv")
S4.contig <- read.csv("~/Documents/GitHub/HNSCC_Descriptive/Data/VDJ_data/Sample4_filtered_contig_annotations.csv")
```


### Combined TCR

Going to go with 3 seperate combinations - unfiltered (combined), remove NA (combined_NA), and how it was originally called (combined.org)

```{r}
library(scRepertoire)
combined <- combineTCR(list(S1.contig, S2.contig, S3.contig, S4.contig), samples = c("S1", "S2", "S3", "S4"), ID = c("US", "US", "S", "S"), cells = "T-AB")

combined_NA <- combineTCR(list(S1.contig, S2.contig, S3.contig, S4.contig), samples = c("S1", "S2", "S3", "S4"), ID = c("US", "US", "S", "S"), cells = "T-AB", removeNA = TRUE)

combined_Org <- combineTCR(list(S1.contig, S2.contig, S3.contig, S4.contig), samples = c("S1", "S2", "S3", "S4"), ID = c("US", "US", "S", "S"), cells = "T-AB", removeNA = TRUE, filterMulti = TRUE)

```

## Loading seurat object

```{r}
library(Seurat)
seurat.combined <- readRDS("~/Documents/GitHub/HNSCC_Descriptive/Data/seurat_combined.rds")
```


### Checking barcodes

```{r}
head(rownames(seurat.combined[[]])) #seurat.combined[[]] grabs the meta.data
head(combined[[3]]$barcode) 
```

### Fixing barcode issue

This has been a consistent issue with the ID variable leading to a mismatch in barcodes compared to a seurat object. Just going to loop the gsub across all three combined objects.

```{r}
combined.list <- list(combined, combined_NA, combined_Org)

for (i in seq_along(combined.list)) {
    combined.list[[i]][[1]][["barcode"]] <- gsub('S1_US', 'S1', combined.list[[i]][[1]][["barcode"]]) #replaces S1_US with S1_
    combined.list[[i]][[2]][["barcode"]] <- gsub('S2_US', 'S2', combined.list[[i]][[2]][["barcode"]])
    combined.list[[i]][[3]][["barcode"]] <- gsub('S3_S', 'S3', combined.list[[i]][[3]][["barcode"]])
    combined.list[[i]][[4]][["barcode"]] <- gsub('S4_S', 'S4', combined.list[[i]][[4]][["barcode"]])
}
```


### Extracting and cleaning meta.data

Looking at the meta data - there is already columns for the TCR information. In addition, as we are going to loop over the 3 combinedTCR objects, I will just extract the meta.data and work with it. I am saving all the vectors from "group" to the end of the data frame.

```{r}
meta.data <- seurat.combined[[]][,which(colnames(meta.data) == "group"):ncol(seurat.combined[[]])]
```

## Merging data

Because we are using the meta data in lieu of the seurat object - I am going to modify the functions of scRepertoire a little and use them.

Here I used the `combineExpression()` and stripped out the parameters - now `combineMeta()` takes the combinedTCR() object, meta.data and the cloneTypes and calculated frequency by condition and the amino acid sequence. 

*Based on reading the code - the assumption I am making here is you want to group frequencies by condition instead of samples* 


```{r}
source("~/Documents/GitHub/HNSCC_Descriptive/R/combineMeta.R")

meta.list <- list()

for (i in seq_along(combined.list)){
    meta.list[[i]] <- combineMeta(combined.list[[i]], meta = meta.data)
}
```

***

# Data visualization

Now we have the meta data from seurat.combined merged with the TCR data we can perform the alluvial comparisons.

## Load Libraries
```{r}
library(ggplot2)
library(ggalluvial)
library(ggfittext)
```

## Chord Diagrams 

Here is a little more inuitive for me, seperate the Stim and US cells, create a matrix that counts the shared clonotype by clusters, correct for overcounting and visuaize both conditions using Chord Diagrams. 

For the most part the function `getCircles()` does all the processing. 

```{r}
library(circlize)

source("~/Documents/GitHub/HNSCC_Descriptive/R/getCircles.R")

Chord.list <- list()
for (i in seq_along(meta.list)) {
    Stim <- subset(meta.list[[i]], condition == "Stim")
    US <- subset(meta.list[[i]], condition == "US")
    Stim <- getCircles(Stim)
    US <- getCircles(US)
    Chord.list[[i]] <- list(Stim, US)
    names(Chord.list[[i]]) <- c("Stim", "US")
}


#Defining grid colors
grid.cols = c(Naive_like_1_CM = "#E41A1C",
                Naive_like_2_SC = "#A6761D",
                Naive_like_3 = "#8DA0CB", 
                Cytotoxic = "#666666", 
                Type_I_IFN = "#A6D854", 
                Stimulated_1= "#984EA3", 
                Stimulated_exhausted = "#1B9E77", 
                Exhausted_1 = "#D95F02", 
                Exhausted_2 = "#7570B3", 
                TRM = "#E7298A", 
                gd_T_g9d2 = "#E6AB02", 
                gd_T_non_g9d2 = "#8DD3C7", 
                MAIT = "#FF7F00", 
                Proliferatice = "#E78AC3")
```

## Plotting Chord Diagrams
```{r}
dir.create("~/Documents/GitHub/HNSCC_Descriptive/Visualizations")
dir <- "~/Documents/GitHub/HNSCC_Descriptive/Visualizations/"

files <- c("nofilter", "NAfilter", "Orgfilter")

#Slight alteration in ciros.track to produce labels that are perpindicular to the circle instead of parallel
for (i in seq_along(Chord.list)) {
    pdf(paste0(dir, files[i], "_Stimulated_TCRcircos.pdf"), height=10, width=10)
chordDiagram(Chord.list[[i]][["Stim"]], grid.col = grid.cols, self.link = 1, annotationTrack = "grid", 
    preAllocateTracks = list(track.height = max(strwidth(unlist(dimnames(Chord.list[[i]][["Stim"]]))))))
circos.track(track.index = 1, panel.fun = function(x, y) {
    xlim = get.cell.meta.data("xlim")
    xplot = get.cell.meta.data("xplot")
    ylim = get.cell.meta.data("ylim")
    sector.name = get.cell.meta.data("sector.index")

    circos.text(mean(xlim), ylim[1], sector.name, facing = "clockwise",
        niceFacing = TRUE, adj = c(0, 0.5)) }, bg.border = NA)
dev.off()

pdf(paste0(dir, files[i], "_Untimulated_TCRcircos.pdf"), height=10, width=10)
chordDiagram(Chord.list[[i]][["US"]], grid.col = grid.cols, self.link = 1, annotationTrack = "grid", 
    preAllocateTracks = list(track.height = max(strwidth(unlist(dimnames(Chord.list[[i]][["US"]]))))))
circos.track(track.index = 1, panel.fun = function(x, y) {
    xlim = get.cell.meta.data("xlim")
    xplot = get.cell.meta.data("xplot")
    ylim = get.cell.meta.data("ylim")
    sector.name = get.cell.meta.data("sector.index")

    circos.text(mean(xlim), ylim[1], sector.name, facing = "clockwise",
        niceFacing = TRUE, adj = c(0, 0.5)) }, bg.border = NA)
dev.off()
}

```

## Alluvial Plots

Still working on this code

```{r}
clust.cols <- c("#E41A1C", # Naive_like_1_CM
                "#A6761D", # Naive_like_2_SC
                "#8DA0CB", # Naive_like_3
                "#666666", # Cytotoxic
                "#A6D854", # Type_I_IFN
                "#984EA3", # Stimulated_1
                "#1B9E77", # Stimulated_exhausted
                "#D95F02", # Exhausted_1
                "#7570B3", # Exhausted_2
                "#E7298A", # TRM
                "#E6AB02", # gd_T_g9d2
                "#8DD3C7", # gd_T_non_g9d2
                "#FF7F00", # MAIT 
                "#E78AC3")

for (i in seq_along(meta.list)) {
graph.df <- meta.list[[i]]
graph.df$barcodes <- rownames(graph.df)

y.axes <- unique(c("condition", "seurat_clusters"))
set.axes <- seq_along(y.axes)
tmp <- graph.df[,c("condition", "seurat_clusters", "CTaa", "barcodes")]
lodes <- to_lodes_form(tmp, key = "x", value = "stratum", 
        id = "alluvium", axes = set.axes, diffuse = c("seurat_clusters", "condition"))

plot <- ggplot(data = lodes, aes(x = x, stratum = stratum, 
                alluvium = alluvium, label = stratum)) +
        geom_stratum() + theme_classic() +
        geom_fit_text(stat = "stratum", infer.label = FALSE, reverse = TRUE) +
        theme(axis.title.x = element_blank(), axis.ticks.x = element_blank()) + 
        geom_flow(aes(fill = lodes[,"seurat_clusters"]), 
            stat = "alluvium", lode.guidance = "forward") + 
        labs(fill = "Cluster") + 
    scale_fill_manual(values = clust.cols) 
ggsave(path = dir, file = paste0(files[i], "_TCRalluvial.pdf"), height=4, width=6)
}

```
