---
title: "04_Manuscript_Figure_4"
author: "Dillon Corvino"
date: "07/03/2022"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

Built with R version `{r} getRversion()`

## Setup {.tabset}

### Dataset information

```{r Dataset_Info}

# HNSCC scRNAseq/scTCRseq dataset
# Cells are HNSCC TILs sorted on Lymphocytes/Live/CD3+/CD4-/CD8+
# Human samples
# Each sample is a pool of 4 patients
# samples 1 and 2 are unstimulated 
# samples 3 and 4 are stimulated 
# Patients in sample 1 match patients in sample 3 and sample 2 pairs with 4
# Data acquired was transcript expression, ADT (antibody expression), and TCRA & B sequences
# This analysis uses just Transcript and TCR data


# General information
# CellRanger was used for sequence alignment/QC/counts/TCR calls - performed by Ross (QIMR, Brisbane, Aus)
# DC was provided the output from CellRanger (counts matrix) and used this as input for analysis within this script

```

### Environment

```{r setup, message = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)

# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory


# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# pipeline variables
long.compute <- FALSE

# load required packages
require(tidyverse)
require(Seurat)
require(useful)
require(ggrepel)
require(clusterProfiler)
require(enrichplot)
require(org.Hs.eg.db)
require(UCell)
require(Nebulosa)
require(ggsci)
library(harmony)
require(irGSEA)
#source("scripts/aimed_analysis/functions.R")



# Create output directory
output.dir <- "results/Manuscript/Figure4/"

if(!dir.exists(paste0(output.dir))){
  dir.create(paste0(output.dir), 
             recursive = TRUE)
}
```




## Type I IFN score HNSCC dataset {.tabset}



### Reading data
```{r reading_data}

# Read in CD8 only dataset
CD8.seurat <- LoadH5Seurat("saves/CD8_seurat.h5Seurat")
Idents(CD8.seurat) <- CD8.seurat@meta.data$Clusters_l1
DefaultAssay(CD8.seurat) <- "RNA"

```


### Define signature
```{r define_signature}

# Compute DEGs and define the Type_I_IFN cluster signature
if(long.compute){
  
  DefaultAssay(CD8.seurat) <- "RNA"
  
  # Calculate DEGs
  IFN.clust.DEGs <- FindMarkers(CD8.seurat,
                                ident.1 = "Type_I_IFN",
                                logfc.threshold = 0,
                                min.pct = 0)
  # Annotate DEGs
  IFN.clust.DEGs <-
    IFN.clust.DEGs%>%
    as_tibble(rownames = "gene_symbol") %>%
    mutate(DE = case_when(
      avg_log2FC > 0 & p_val_adj <= 0.1 ~ "up",
      avg_log2FC < 0 & p_val_adj <= 0.1 ~ "down",
      TRUE ~ "notDE"))
  
  # Export results
  write_csv(x = IFN.clust.DEGs,
            file = paste0(output.dir, "IFN_clust_DEGs.csv"))
  
  
  # Define signature as top DEGs only
  IFN.sig <- IFN.clust.DEGs %>%
    filter(p_val_adj <= 0.05) %>%
    top_n(20, avg_log2FC)
  
  
  # Export results
  write_csv(x = IFN.sig,
            file = paste0(output.dir, "IFN_signature.csv"))
  
  
}


# create signatures
IFN.clust.DEGs <- read.csv(paste0(output.dir, "IFN_clust_DEGs.csv"))

Sig.10 <- IFN.clust.DEGs %>%
  filter(p_val_adj <= 0.05) %>%
  top_n(10, avg_log2FC) %>%
  pull(gene_symbol)

Sig.10.down <- IFN.clust.DEGs %>%
  filter(p_val_adj <= 0.05) %>%
  top_n(-10, avg_log2FC) %>%
  pull(gene_symbol)

Sig.10.down <- paste0(Sig.10.down, "-")
Sig.10 <- paste0(Sig.10, "+")



Signature.list <- list(Top_10 = Sig.10,
                       Top10_bot10 = c(Sig.10, Sig.10.down))

signature.names <- paste0(names(Signature.list), "_UCell")




# extract hallmark pathways
# x <- msigdbr::msigdbr(species = "Homo sapiens",
#                       category = "H")
# 
# Hallmark.IFNa.response <- x[x$gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE", ]$gene_symbol
# Hallmark.IFNa.response <- unique(Hallmark.IFNa.response)
# 
# 
# Hallmark.IFNg.response <- x[x$gs_name == "HALLMARK_INTERFERON_GAMMA_RESPONSE", ]$gene_symbol
# Hallmark.IFNg.response <- unique(Hallmark.IFNg.response)
# 
# 
# # create list of signatures
# Signature.list <- list(Positive_sig = IFN.sig, 
#                        IFNa_Hallmark = Hallmark.IFNa.response,
#                        IFNg_Hallmark = Hallmark.IFNg.response)
# 
# signature.names <- paste0(names(Signature.list), "_UCell")

```


### UCell HNSCC dataset
```{r Ucell_HNSCC_dataset}

# Calculate Ucell scores for signatures
DefaultAssay(CD8.seurat) <- "RNA"
CD8.seurat <- AddModuleScore_UCell(CD8.seurat,
                                     features = Signature.list,
                                     ncores = 4)



# Visualise results
VlnPlot(CD8.seurat,
        pt.size = 0, 
        cols = CD8.cols,
        same.y.lims = T,
        features = signature.names[1],
        group.by = "Clusters_l1") + 
  NoLegend()

dev.copy(pdf, paste0(output.dir, "HNSCC_VlnPlot_UCell_scores_top10.pdf"))
dev.off()


 
Plot_Density_Custom(CD8.seurat, 
                    features =  signature.names[1],
                    custom_palette = batlow.pal,
                    joint = FALSE, 
                    pt.size = 1, 
                    reduction = "umap")

dev.copy(pdf, paste0(output.dir, "HNSCC_Nebulosa_top10.pdf"))
dev.off()



```

### UCell HNSCC dataset noIFN cluster
```{r Ucell_HNSCC_dataset}

# Calculate Ucell scores for signatures
DefaultAssay(CD8.seurat) <- "RNA"
keep.var <- levels(Idents(CD8.seurat))
keep.var <- keep.var[-5]

temp.seurat <- subset(CD8.seurat, idents = keep.var)
levels(Idents(temp.seurat))

temp.seurat <- AddModuleScore_UCell(temp.seurat,
                                     features = Signature.list,
                                     ncores = 4)


# Visualise results
VlnPlot(temp.seurat,
        pt.size = 0, 
        cols = CD8.cols,
        same.y.lims = T,
        features = signature.names[1],
        group.by = "Clusters_l1") + 
  NoLegend()

dev.copy(pdf, paste0(output.dir, "HNSCC_VlnPlot_UCell_scores_top10_no_IFN.pdf"))
dev.off()

 

Plot_Density_Custom(temp.seurat, 
                    features =  signature.names[1],
                    custom_palette = batlow.pal,
                    joint = FALSE, 
                    pt.size = 1, 
                    reduction = "umap")

dev.copy(pdf, paste0(output.dir, "HNSCC_Nebulosa_top10_no_IFN.pdf"))
dev.off()


dev.copy(pdf, paste0(output.dir, "HNSCC_Nebulosa_top10_no_IFN.pdf"))
dev.off()

rm(temp.seurat)


```

## Type I IFN score Cillo dataset {.tabset}

### read in cillo data

```{r read_dataset_cillo}

Cillo.CD8.Seurat <- LoadH5Seurat("data/external_datasets/Cillo_TIL_CD8_only.h5Seurat")


head(Cillo.CD8.Seurat@meta.data)


DimPlot(Cillo.CD8.Seurat, 
        reduction = "tsne", 
        group.by = "Pt_ID")



```
 
### UCell Cillo
```{r cillo_ucell}

# Calculate Ucell scores for signatures
DefaultAssay(Cillo.CD8.Seurat) <- "RNA"
Cillo.CD8.Seurat <- AddModuleScore_UCell(Cillo.CD8.Seurat,
                                     features = Signature.list,
                                     ncores = 4)

# Visualise results
VlnPlot(Cillo.CD8.Seurat,
        pt.size = 0, 
        same.y.lims = T,
        features = signature.names[1],
        group.by = "cd8_cells_clusterID") + 
  NoLegend()

dev.copy(pdf, paste0(output.dir, "Cillo_VlnPlot_UCell_scores_top10.pdf"))
dev.off()


Plot_Density_Custom(Cillo.CD8.Seurat, 
                    features =  signature.names[1],
                    custom_palette = batlow.pal,
                    joint = FALSE, 
                    pt.size = 1, 
                    reduction = "umap")

dev.copy(pdf, paste0(output.dir, "Cillo_Nebulosa_top10.pdf"))
dev.off()


# UMAPPlot(Cillo.CD8.Seurat)
# 
# 
# Idents(Cillo.CD8.Seurat) <- Cillo.CD8.Seurat@meta.data$cd8_cells_clusterID 
# Cluster_Highlight_Plot(seurat_object = Cillo.CD8.Seurat,
#                        cluster_name = "2",
#                        reduction = "umap",
#                        highlight_color = "navy",
#     background_color = "lightgray")
# 


```

## Type I IFN score pancancer dataset {.tabset}

### Prepare dataset / read and filter for CD8

```{r read_dataset_pancancer}

if(long.compute){
pancancer.seurat <- LoadH5Seurat("data/external_datasets/Utility_dataset.h5Seurat")

Idents(pancancer.seurat) <- pancancer.seurat@meta.data$consensus.Tcell
pancancer.cd8 <- subset(pancancer.seurat, idents = "CD8")

# 1 cell of mixed annoation and 2 other cells remain - remove these
Idents(pancancer.cd8) <- pancancer.cd8@meta.data$consensus.major
pancancer.cd8 <- subset(pancancer.cd8, idents = "T_cell")

SaveH5Seurat(pancancer.cd8, 
             "data/external_datasets/pancancer_cd8.h5Seurat")

# 
Idents(pancancer.cd8) <- pancancer.cd8@meta.data$Type
pancancer.cd8 <- subset(pancancer.cd8, idents = c("Tumor", "Normal"))

SaveH5Seurat(pancancer.cd8, 
             "data/external_datasets/pancancer_cd8_tumor_normal.h5Seurat")


# Normalise and harmonize dataset and calculate umap projection
pancancer.cd8 <- 
  pancancer.cd8 %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  harmony::RunHarmony(reduction = "pca",
                      group.by.vars = c("Cohort"), 
                      plot_convergence = TRUE, 
                      verbose = TRUE) %>%
  RunUMAP(reduction = "harmony", 
          dims = 1:20) %>%
  FindNeighbors(reduction = "harmony",
                dims = 1:20) %>%
  FindClusters(resolution = 0.5)

# Export harmony corrected object
SaveH5Seurat(pancancer.cd8, 
             "data/external_datasets/pancancer_cd8_tumor_normal_harmonised.h5Seurat", 
             overwrite = TRUE)

}

```
 
### apply ucell score to pancancer

```{r ucell_pancancer}


if(long.compute){

pancancer.cd8 <- LoadH5Seurat("data/external_datasets/pancancer_cd8_tumor_normal_harmonised.h5Seurat")

# Calculate Ucell scores for signatures
DefaultAssay(pancancer.cd8) <- "RNA"
pancancer.cd8 <- UCell::AddModuleScore_UCell(pancancer.cd8,
                                             features = Signature.list,
                                             ncores = 4)

# Export harmony corrected object
SaveH5Seurat(pancancer.cd8, 
             "data/external_datasets/pancancer_cd8_tumor_normal_harmonised.h5Seurat", 
             overwrite = TRUE)

}


# Visualise results
VlnPlot(pancancer.cd8,
        pt.size = 0, 
        same.y.lims = T,
        features = signature.names[1],
        group.by = "seurat_clusters") + 
  NoLegend()

dev.copy(pdf, paste0(output.dir, "Pancancer_VlnPlot_UCell_scores_top10.pdf"))
dev.off()


 
Plot_Density_Custom(pancancer.cd8, 
                    features =  signature.names[1],
                    custom_palette = batlow.pal,
                    joint = FALSE, 
                    pt.size = 1, 
                    reduction = "umap")

dev.copy(pdf, paste0(output.dir, "pancancer_Nebulosa_top10.pdf"))
dev.off()








# extract normed values 

var.by.cluster <- table(pancancer.cd8@meta.data$Type,
                        pancancer.cd8@meta.data$seurat_clusters)

total.cells.type <- rowSums(var.by.cluster)

output.df <- var.by.cluster*NA

# normalise along row
for(i in 1:length(total.cells.type)){
output.df[i,] <- (var.by.cluster[i,]/total.cells.type[i]) * 100
}


output.df[,8]

# Plot freq of IFN cluster
barplot(output.df[,8], 
        las = 2,
        col = c("gray", "blue"), 
        cex.names = 0.6, 
        ylim = c(0, 5))

dev.copy(pdf, paste0(output.dir, "pancancer_barplot_cluster_7_normed_type.pdf"))
dev.off()

write.csv(output.df, paste0(output.dir, "pancancer_barplot_cluster_7_normed_by_type.csv"))






var.by.cluster <- table(pancancer.cd8@meta.data$Tissue,
                        pancancer.cd8@meta.data$seurat_clusters)

total.cells <- rowSums(var.by.cluster)

output.df <- var.by.cluster*NA

# normalise along row
for(i in 1:length(total.cells)){
output.df[i,] <- (var.by.cluster[i,]/total.cells[i]) * 100
}


output.df[,8]

# Plot freq of IFN cluster
barplot(output.df[,8], 
        las = 2,
        col = c("gray", "blue"), 
        cex.names = 0.6, 
        ylim = c(0, 30))

dev.copy(pdf, paste0(output.dir, "pancancer_barplot_cluster_7_normed_by_tissue.pdf"))
dev.off()

write.csv(output.df, paste0(output.dir, "pancancer_barplot_cluster_7_normed_by_tissue.csv"))

























x <- table(pancancer.cd8@meta.data$Tissue)
min(x)

Idents(pancancer.cd8) <- pancancer.cd8@meta.data$Tissue
temp.seurat <- subset(pancancer.cd8, downsample = 700)

Idents(temp.seurat) <- temp.seurat@meta.data$seurat_clusters

scCustomize::DimPlot_scCustom(temp.seurat, 
                              pt.size = 1,
                              num_columns = 4,
                              split.by = "Tissue")

dev.copy(pdf, paste0(output.dir, "pancancer_umap_by_tissue_downsampled.pdf"))
dev.off()




```


