---
title: "HNSCC_scRNAseq_Descriptive_paper"
author: "Dillon Corvino"
date: "03/02/2020"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# HNSCC scRNAseq/scTCRseq dataset
# Cells are HNSCC TILs sorted on Lymphocytes/Live/CD3+/CD4-/CD8+
# Human samples
# Each sample is a pool of 4 patients
# samples 1 and 2 are unstimulated 
# samples 3 and 4 are stimulated 
# Patients in sample 1 match patients in sample 3 and sample 2 pairs with 4
# Data acquired was transcript expression, ADT (antibody expression), and TCRA & B sequences
# This analysis uses just Transcript and TCR data


# General information
# CellRanger was used for sequence alignment/QC/counts/TCR calls - performed by Ross (QIMR, Brisbane, Aus)
# DC was provided the output from CellRanger (counts matrix) and used this as input for analysis within this script

```

### Environment
```{r Environment_setup, message = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
start.time <- Sys.time()
quick.load <- TRUE
long.compute <- FALSE

###################
# Install packages
###################

# CITEseq requires latest version of Seurat i.e at least V3
# To install the development version of seurat
# devtools::install_github(repo = 'satijalab/seurat', ref = 'develop')

pkgs <- c("remedy", "Seurat", "dplyr", "rstudioapi",
          "cowplot", "ggplot2", "grid", "gridExtra",
          "styler", "stringr", "inlmisc", "RColorBrewer",
          "readxl", "devtools", "tidyverse", "hdf5r", "scales", "useful")

for(i in 1:length(pkgs)){
  if(!require(pkgs[i], character.only = T)){
    install.packages(pkgs[i])
    require(pkgs[i], character.only = T)
  }else{
    require(pkgs[i], character.only = T)
  }
}

pkgs <- c("gplots", "fgsea", "biomaRt", "clusterProfiler", 
          "GSEABase", "org.Hs.eg.db", "org.Mm.eg.db", "pcaMethods",
          "SingleCellExperiment", "batchelor", "DelayedArray", "DelayedMatrixStats",
          "limma", "SummarizedExperiment")

for(i in 1:length(pkgs)){
  if(!require(pkgs[i], character.only = T)){
    BiocManager::install(pkgs[i])
    require(pkgs[i], character.only = T)
  }else{
    require(pkgs[i], character.only = T)
  }
}


#####################
# Github packages
#####################

# Installed packages 

# library("devtools")

# usethis::browse_github_pat()
# usethis::edit_r_environ()
# GITHUB_PAT = "d8207153aef7b295cdf66eb1e1b2a2ed38b0ca18"
# R_MAX_VSIZE = 30Gb

# celltalker
#devtools::install_github("arc85/celltalker")

# Scillus 
#devtools::install_github("xmc811/Scillus", ref = "development")

#devtools::install_github('satijalab/seurat-data')

#devtools::install_github("velocyto-team/velocyto.R")

#devtools::install_github("dynverse/dyno")

#devtools::install_github('cole-trapnell-lab/leidenbase')
#devtools::install_github('cole-trapnell-lab/monocle3')

#devtools::install_github("immunomind/immunarch")
#devtools::install_github("Coolgenome/iTALK", build_vignettes = TRUE)

#########################
# Failed installations
#########################

# I have been getting a killed:9 error installing the following github packages. 
# I believe the issue is clang or something to do with the compiler. Unable to fix the issue. Many open github issues on this error

# For using scRep package I have made a work around to get access to the functions
#devtools::install_github("ncborcherding/scRepertoire")

#devtools::install_github('satijalab/seurat-wrappers')


# Moncle packages to load 
#library(monocle3)
#library(Seurat)
#library(SeuratData)
#library(SeuratWrappers)
#library(ggplot2)
#library(patchwork)
#library(magrittr)


#.rs.restartR()


# Failed
#remotes::install_git("https://gitlab.unil.ch/carmona/ProjecTILs.git", ref = "v0.4.1")

# TILPRED install works, but only currently for mouse datasets
#remotes::install_github("carmonalab/TILPRED")


####################
# Colour scheme
####################

Condition.cols <- c("turquoise", "red")

clust.cols <- c("#E41A1C", # Naive_like_1_CM
                "#A6761D", # Naive_like_2_SC
                "#8DA0CB", # Naive_like_3
                "#666666", # Cytotoxic
                "#A6D854", # Type_I_IFN
                "#984EA3", # Stimulated_1
                "#1B9E77", # Stimulated_exhausted
                "#D95F02", # Exhausted_1
                "#7570B3", # Exhausted_2
                "#E7298A", # TRM
                "#E6AB02", # gd_T_g9d2
                "#8DD3C7", # gd_T_non_g9d2
                "#FF7F00", # MAIT 
                "#E78AC3") # Proliferative


module.cols <- c("#E41A1C", # Naive_Stem
                 "#1B9E77", # Activated
                 "#7570B3", # Exhaustion
                 "#FF7F00", # Innate
                 "#E78AC3") # Cycling

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
working.dir <- getwd()

# create output directories
if(!dir.exists("Exported_RDS_files")){dir.create("Exported_RDS_files", recursive = T)}
if(!dir.exists("output")){dir.create("output", recursive = T)}
if(!dir.exists("output/figures")){dir.create("output/figures", recursive = T)}
if(!dir.exists("output/tables")){dir.create("output/tables", recursive = T)}
if(!dir.exists("output/QC")){dir.create("output/QC", recursive = T)}


# load saved seurat object with cluster annotations and imputation performed
if(quick.load){
  seurat.combined <- readRDS("Exported_RDS_files/seurat_combined.rds")
}


```

## TCR analysis - formatting {.tabset}

### Load scRepertoire package
```{r Load_scRep_Package}

# If using R < 4, should use the dev version with @dev
#devtools::install_github("ncborcherding/scRepertoire")
#install.packages("ggalluvial")

# Package fails to load - use this trick to get tools of the package 
devtools::load_all("~/Documents/Work/Sciebo/Scripts/Packages/scRepertoire/")


```
 
### Format data with scRepertoire
```{r Format_data_with_scRep}
#################################
# Read in and format data
#################################

# read in contig files
S1.contig <- read.csv("Input_data/VDJ_data/Sample1_filtered_contig_annotations.csv", stringsAsFactors = FALSE)
S2.contig <- read.csv("Input_data/VDJ_data/Sample2_filtered_contig_annotations.csv", stringsAsFactors = FALSE)
S3.contig <- read.csv("Input_data/VDJ_data/Sample3_filtered_contig_annotations.csv", stringsAsFactors = FALSE)
S4.contig <- read.csv("Input_data/VDJ_data/Sample4_filtered_contig_annotations.csv", stringsAsFactors = FALSE)

# create list of contigs
contig.list <- list(S1.contig, S2.contig,
                    S3.contig, S4.contig)

# Combine contig list
combined.contig <- combineTCR(contig.list, 
                              samples = c("S1", "S2", "S3", "S4"), 
                              ID = c("US", "US", "Stim", "Stim"), 
                              cells = "T-AB",
                              removeNA = TRUE,
                              removeMulti = FALSE, 
                              filterMulti = TRUE) # take the top 2 expressed chains for cells with multiple chains 

# on average, filteringMulti results in approx. 200 more cells for each sample than outright removing cells


# Need to amend BarcodeID to match with seurat object
combined.contig$S1_US$barcode <- gsub("US_", "", combined.contig$S1_US$barcode)
combined.contig$S2_US$barcode <- gsub("US_", "", combined.contig$S2_US$barcode)
combined.contig$S3_Stim$barcode <- gsub("Stim_", "", combined.contig$S3_Stim$barcode)
combined.contig$S4_Stim$barcode <- gsub("Stim_", "", combined.contig$S4_Stim$barcode)


# Combine seurat object with contig info
seurat.tcr <- combineExpression(combined.contig,
                                seurat.combined,
                                cloneCall = "aa",
                                cloneTypes = c(None = 0, Single = 1, Small = 5, Medium = 10, Large = 20, Hyperexpanded = 150),
                                groupBy = "none", # Alternative frequencies are calculated in next code Chunk
                                filterNA = TRUE) # should seurat object be subsetted to remove NA values

# Max freq of a clonotype is 127, therefore reset cloneTypes accordingly

# Seurat object went from 11658 samples to 6625 samples 
# Therefore 56% of original dataset remains 


# Remove frequency column calculated by combineExpression function
logic.vec <- grepl("Frequency", colnames(seurat.tcr@meta.data))
seurat.tcr@meta.data[ ,logic.vec] <- NULL
head(seurat.tcr@meta.data)

# Ensure seurat object cloneType var is a factor with correct levels
seurat.tcr@meta.data$cloneType <- factor(seurat.tcr@meta.data$cloneType, 
                                         levels = c("Hyperexpanded (20 < X <= 150)",
                                                    "Large (10 < X <= 20)", 
                                                    "Medium (5 < X <= 10)",
                                                    "Small (1 < X <= 5)", 
                                                    "Single (0 < X <= 1)"))

# remove unneeded objects
rm(combined.contig, contig.list, S1.contig, S2.contig, S3.contig, S4.contig)

```

### Count clonotype frequencies and normalise
```{r Count_and_normalise_clonotype_frequencies}

####################################################################################
# Need to count clonotype occurrence per condition and then normalise this value 
####################################################################################

# create temp meta.data df and append cell_ID as column (lost in dplyr workflow)
meta.data.temp <- seurat.tcr@meta.data
meta.data.temp$cell_ID <- rownames(seurat.tcr@meta.data)

corner(meta.data.temp, c = ncol(meta.data.temp))


# Get total number of cells in each condition
US.cell.n <- table(meta.data.temp$condition)[1] # 3,184 
Stim.cell.n <- table(meta.data.temp$condition)[2] # 3,441
total.cell.n <- US.cell.n + Stim.cell.n # 6,625 

#dim(meta.data.temp) // sanity check of total cell number




####################################################################################
# Count occurence of clonotypes across entire dataset 
####################################################################################
# Also: Normalise n() by # of cells within dataset
meta.data.temp <- meta.data.temp %>%
  dplyr::group_by(CTaa) %>%
  dplyr::mutate(Total_clonotype_n = n()) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Total_clonotype_freq = (Total_clonotype_n/total.cell.n)*100) %>%
  dplyr::ungroup()

####################################################################################
# Count occurence of clonotypes across entire condition 
####################################################################################
# Also: Normalise n() by # of cells within condition
meta.data.temp <- meta.data.temp %>%
  dplyr::group_by(CTaa, condition) %>%
  dplyr::mutate(Clone_conditionwise_n = n()) %>%
  dplyr::group_by(condition) %>%
  dplyr::mutate(Clone_conditionwise_freq = case_when(condition == "US" ~ (Clone_conditionwise_n/US.cell.n)*100, 
                                                         condition == "Stim" ~ (Clone_conditionwise_n/Stim.cell.n)*100)) %>%
  dplyr::ungroup()

####################################################################################
# Count occurence of clonotypes within each cluster per condition 
####################################################################################
# Also: Normalise n() by # of cells within condition
meta.data.temp <- meta.data.temp %>%
  dplyr::group_by(CTaa, condition, seurat_clusters) %>%
  dplyr::mutate(Clone_Clust_conditionwise_n = n()) %>%
  dplyr::group_by(condition) %>%
  dplyr::mutate(Clone_Clust_conditionwise_freq = case_when(condition == "US" ~ (Clone_Clust_conditionwise_n/US.cell.n)*100, 
                                                         condition == "Stim" ~ (Clone_Clust_conditionwise_n/Stim.cell.n)*100)) %>%
  dplyr::ungroup()


######################################
# Add new metadata to seurat object
######################################

#########
# Total 
#########

# Add total clonotype n value
seurat.tcr <- AddMetaData(object = seurat.tcr, 
                          col.name = "Total_clonotype_n",
                          metadata = meta.data.temp$Total_clonotype_n)

# Add total clonotype freq
seurat.tcr <- AddMetaData(object = seurat.tcr, 
                          col.name = "Total_clonotype_freq",
                          metadata = meta.data.temp$Total_clonotype_freq)

##################
# Condition-wise 
##################

# Add clone condition-wise n value
seurat.tcr <- AddMetaData(object = seurat.tcr, 
                          col.name = "Clone_conditionwise_n",
                          metadata = meta.data.temp$Clone_conditionwise_n)

# Add Clone condition-wise freq
seurat.tcr <- AddMetaData(object = seurat.tcr, 
                          col.name = "Clone_conditionwise_freq",
                          metadata = meta.data.temp$Clone_conditionwise_freq)

################################
# Cluster and Condition-wise 
################################

# Add Clone Clust and condition-wise n value
seurat.tcr <- AddMetaData(object = seurat.tcr, 
                          col.name = "Clone_Clust_conditionwise_n",
                          metadata = meta.data.temp$Clone_Clust_conditionwise_n)

# Add Clone Clust and condition-wise Freq
seurat.tcr <- AddMetaData(object = seurat.tcr, 
                          col.name = "Clone_Clust_conditionwise_freq",
                          metadata = meta.data.temp$Clone_Clust_conditionwise_freq)


#############################################
# Scale variables for downstream plotting
#############################################
scale.var <- c(0, 10)

# Scale total clonotype number
seurat.tcr@meta.data$Total_clonotype_scaled <- scales::rescale(seurat.tcr@meta.data$Total_clonotype_n, 
                                                               to = scale.var)


# Scale clonotype per condition freq
seurat.tcr@meta.data$Clone_conditionwise_scaled <- scales::rescale(seurat.tcr@meta.data$Clone_conditionwise_freq, 
                                                                   to = scale.var)

# Scale clonotype per cluster per condition freq
seurat.tcr@meta.data$Clone_Clust_conditionwise_scaled <- scales::rescale(seurat.tcr@meta.data$Clone_Clust_conditionwise_freq, 
                                                                         to = scale.var)




rm(meta.data.temp)


```
 
### Export or read in tcr seurat object
```{r export_RDS}

if(quick.load){
  seurat.tcr <- readRDS("Exported_RDS_files/seurat_tcr.rds")
}else if(!quick.load){
  saveRDS(seurat.tcr, file = "Exported_RDS_files/seurat_tcr.rds")
}

```

## TCR analysis - General analysis {.tabset}

### Heatmaps VJ gene usage
```{r V_J_usage_heatmaps}

# Move to working dir
setwd(working.dir)

# Create output directories
if(!dir.exists("output/figures/TCRseq/VJ_usage")){
  dir.create("output/figures/TCRseq/VJ_usage",
             recursive = T)}

# set and move to output.dir
output.dir <- "output/figures/TCRseq/VJ_usage"
setwd(output.dir)



########################
# Heatmap functions
########################

hclust.func <- function(x){
  hclust(x, method = 'ward.D2')
}

dist.func <- function(x){
  dist(x,method = 'binary')
}

col.palette <- colorRampPalette(brewer.pal(8, "RdYlBu"))(600)

######################################################
# Plot V/J gene usage as a log2() of cell number
######################################################

# TRA-V gene usage

x <- table(seurat.tcr@meta.data$v_gene_A, seurat.tcr@meta.data$seurat_clusters)

cells.remove <- c("gd_T_non_g9d2", "gd_T_g9d2", "MAIT")

x <- x[ ,-grep(paste0(cells.remove, collapse = "|"), colnames(x))]


heatmap.2(as.matrix(log2(x+1)), 
          scale = "none",
          trace = "none",
          Colv = TRUE,
          cexRow = 0.5, 
          margins = c(8, 5),
          cexCol = 0.8, 
          distfun = dist.func,
          hclustfun = hclust.func,
          srtCol = 90,
          col = rev(col.palette))     

dev.copy(pdf, "Heatmap_TRA_V_usage.pdf")
dev.off()

# TRA-J gene usage

x <- table(seurat.tcr@meta.data$j_gene_A, seurat.tcr@meta.data$seurat_clusters)

cells.remove <- c("gd_T_non_g9d2", "gd_T_g9d2", "MAIT")

x <- x[ ,-grep(paste0(cells.remove, collapse = "|"), colnames(x))]


heatmap.2(as.matrix(log2(x+1)), 
          scale = "none",
          trace = "none",
          Colv = TRUE,
          cexRow = 0.5, 
          margins = c(8, 5),
          cexCol = 0.8, 
          distfun = dist.func,
          hclustfun = hclust.func,
          srtCol = 90,
          col = rev(col.palette))     

dev.copy(pdf, "Heatmap_TRA_V_usage.pdf")
dev.off()


# TRB-V gene usage

x <- table(seurat.tcr@meta.data$v_gene_B, seurat.tcr@meta.data$seurat_clusters)

cells.remove <- c("gd_T_non_g9d2", "gd_T_g9d2", "MAIT")

x <- x[ ,-grep(paste0(cells.remove, collapse = "|"), colnames(x))]


heatmap.2(as.matrix(log2(x+1)), 
          scale = "none",
          trace = "none",
          Colv = TRUE,
          cexRow = 0.5, 
          margins = c(8, 5),
          cexCol = 0.8, 
          distfun = dist.func,
          hclustfun = hclust.func,
          srtCol = 90,
          col = rev(col.palette))     

dev.copy(pdf, "Heatmap_TRB_V_usage.pdf")
dev.off()



# TRB-J gene usage

x <- table(seurat.tcr@meta.data$j_gene_B, seurat.tcr@meta.data$seurat_clusters)

cells.remove <- c("gd_T_non_g9d2", "gd_T_g9d2", "MAIT")

x <- x[ ,-grep(paste0(cells.remove, collapse = "|"), colnames(x))]


heatmap.2(as.matrix(log2(x+1)), 
          scale = "none",
          trace = "none",
          Colv = TRUE,
          cexRow = 0.5, 
          margins = c(8, 5),
          cexCol = 0.8, 
          distfun = dist.func,
          hclustfun = hclust.func,
          srtCol = 90,
          col = rev(col.palette))     

dev.copy(pdf, "Heatmap_TRB_J_usage.pdf")
dev.off()




# Reset working dir
setwd(working.dir)




```

### Heatmap of top clonotypes
```{r Heatmap_top_clonotypes}

# Move to working dir
setwd(working.dir)

# Create output directories
if(!dir.exists("output/figures/TCRseq/Top_clonotypes_heatmaps")){
  dir.create("output/figures/TCRseq/Top_clonotypes_heatmaps",
             recursive = T)}

# set and move to output.dir
output.dir <- "output/figures/TCRseq/Top_clonotypes_heatmaps"
setwd(output.dir)



########################
# Heatmap functions
########################

hclust.func <- function(x){
  hclust(x, method = 'ward.D2')
}

dist.func <- function(x){
  dist(x,method = 'binary')
}

col.palette <- colorRampPalette(brewer.pal(8, "RdYlBu"))(600)












# Get data of how many cells in each cluster are assigned to what clonotypes
clonotype.data <- table(seurat.tcr@meta.data$CTaa, seurat.tcr@meta.data$seurat_clusters)

# remove non-T cell clusters
cells.remove <- c("gd_T_non_g9d2", "gd_T_g9d2")
cells.remove.index <- grep(paste0(cells.remove, collapse = "|"), colnames(clonotype.data))
clonotype.data <- clonotype.data[ ,-cells.remove.index]

# also filter colouring variable to remove non-T cell clusters
temp.cols <- clust.cols[-cells.remove.index]






# Plot the top 10 clonotypes
aa.keep <- sort(rowSums(clonotype.data), decreasing = TRUE)[1:10]

keep.vec <- rownames(clonotype.data) %in% names(aa.keep)

plot.data <- clonotype.data[keep.vec, ]


heatmap.2(as.matrix(log2(plot.data+1)), 
          scale = "none",
          trace = "none",
          ColSideColors = temp.cols,
          Colv = TRUE,
          cexRow = 0.8, 
          margins = c(8, 15),
          cexCol = 0.8, 
          distfun = dist.func,
          hclustfun = hclust.func,
          srtCol = 90,
          col = rev(col.palette))     

dev.copy(pdf, "Heatmap_top_10_clonotypes.pdf")
dev.off()



# Plot the top 50 clonotypes
aa.keep <- sort(rowSums(clonotype.data), decreasing = TRUE)[1:50]

keep.vec <- rownames(clonotype.data) %in% names(aa.keep)

plot.data <- clonotype.data[keep.vec, ]


heatmap.2(as.matrix(log2(plot.data+1)), 
          scale = "none",
          trace = "none",
          ColSideColors = temp.cols,
          Colv = TRUE,
          cexRow = 0.5, 
          labRow = FALSE,
          margins = c(8, 5),
          cexCol = 0.8, 
          distfun = dist.func,
          hclustfun = hclust.func,
          srtCol = 90,
          col = rev(col.palette))     

dev.copy(pdf, "Heatmap_top_50_clonotypes.pdf")
dev.off()



# Plot the all clonotypes in > 3 cells

aa.keep <- rowSums(clonotype.data) > 3

keep.vec <- rownames(clonotype.data) %in% names(aa.keep)[aa.keep]

plot.data <- clonotype.data[keep.vec, ]

heatmap.2(as.matrix(log2(plot.data+1)), 
          scale = "none",
          trace = "none",
          ColSideColors = temp.cols,
          Colv = TRUE,
          cexRow = 0.5, 
          labRow = FALSE,
          margins = c(8, 5),
          cexCol = 0.8, 
          distfun = dist.func,
          hclustfun = hclust.func,
          srtCol = 90,
          col = rev(col.palette))     

dev.copy(pdf, "Heatmap_top_all_clonotypes_in_morethan_3_cells.pdf")
dev.off()

# Reset working dir
setwd(working.dir)
```

### Analysis with scRepertoire
```{r scRepertoire_analysis}

# Ensure dir back in working
setwd(working.dir)

# Create output directories
if(!dir.exists("output/figures/TCRseq/scRepertoire/clusters")){
  dir.create("output/figures/TCRseq/scRepertoire/clusters",
             recursive = T)}

if(!dir.exists("output/figures/TCRseq/scRepertoire/modules")){
  dir.create("output/figures/TCRseq/scRepertoire/modules",
             recursive = T)}

output.dir <- "output/figures/TCRseq/scRepertoire"
setwd(output.dir)

#########################
# Set uniform variables 
#########################

# for clonalProportion split
split.var <- c(1, 5, 10, 20, 50, 500, 5000, 1e+05)

# for clonalHomeostasis split
homeostasis.var <- c(Rare = 0.001, 
                     very_small = 0.002,
                     Small = 0.005, 
                     Medium = 0.01, 
                     Large = 0.05, 
                     very_Large = 0.1, 
                     Hyperexpanded = 1)


#########################################################
# Perform basic clonotype analysis and visualisation
#########################################################

# quantify % unique clonotype per sample
quantContig(combined.contig,
            cloneCall="aa", 
            scale = TRUE)

dev.copy(pdf, "Quant_contig_sample.pdf")
dev.off()

# quantify % unique clonotype per group
quantContig(combined.contig,
            cloneCall="aa", 
            group = "ID",
            scale = TRUE)

dev.copy(pdf, "Quant_contig_treatment.pdf")
dev.off()


# abundance of clonotypes
abundanceContig(combined.contig,
                cloneCall = "aa",
                group = "ID", 
                scale = TRUE)

dev.copy(pdf, "Clonotype_abundance_by_treatment.pdf")
dev.off()



# Chain length overview 
lengthContig(combined.contig, 
             cloneCall="aa", 
             chains = "combined",
             group = "ID") 

dev.copy(pdf, "Chain_length_combined.pdf")
dev.off()

# Chain length by chain
lengthContig(combined.contig, 
             cloneCall="aa", 
             chains = "single",
             group = "ID")

dev.copy(pdf, "Chain_length_by_locus.pdf")
dev.off()


## Something looks weird with the alluvial plots - lines dont line up

# Compare clonotypes across samples

compareClonotypes(combined.contig,
                  numbers = 20, 
                  samples = c("S1_US", "S3_Stim"), 
                  cloneCall = "aa", 
                  graph = "alluvial") + NoLegend()


compareClonotypes(combined.contig,
                  numbers = 20, 
                  samples = c("S2_US", "S4_Stim"), 
                  cloneCall = "aa", 
                  graph = "alluvial") + NoLegend()



# Overview of clonal homeostasis
clonalHomeostasis(combined.contig, 
                  cloneType = homeostasis.var,
                  cloneCall = "aa")

dev.copy(pdf, "clonalHomeostasis.pdf")
dev.off()


# Clonal Proportion
clonalProportion(combined.contig,
                 split = split.var,
                 cloneCall = "aa") 

dev.copy(pdf, "clonalProportion.pdf")
dev.off()

# clonal overlap
clonalOverlap(combined.contig, 
              cloneCall = "aa", 
              method = "morisita")

dev.copy(pdf, "cloncalOverlap.pdf")
dev.off()


# clone size distribution
clonesizeDistribution(combined.contig, 
                      cloneCall = "aa", 
                      method = "ward.D2")

dev.copy(pdf, "cloneSizeDistribution.pdf")
dev.off()

# Clonal Diversity 
clonalDiversity(combined.contig,
                cloneCall = "aa",
                group = "ID")

dev.copy(pdf, "clonalDiversity.pdf")
dev.off()



########################################################################
# perform basic repertoire analysis across clusters & modules
########################################################################

# Modified function, original was not subsetting correctly
expression2List.mod <- function(sc, group){
  meta <- grabMeta(sc)
  unique.var <- str_sort(as.character(unique(meta[,group])), numeric = TRUE)
  df <- NULL
  for (i in seq_along(unique.var)) {
    subset.var <- meta[meta[,group] == unique.var[i],]
    df[[i]] <- subset.var
  }
  names(df) <- unique.var
  
  return(df)
}


tcr.clusters <- expression2List.mod(seurat.tcr, group = "seurat_clusters")

tcr.modules <- expression2List.mod(seurat.tcr, group = "Module")



####################
# Clonal Diversity 
####################

# Clusters
clonalDiversity(tcr.clusters, 
                cloneCall = "aa")

dev.copy(pdf, "clusters/clonalDiversity_clusters.pdf")
dev.off()

# Modules
clonalDiversity(tcr.modules, 
                cloneCall = "aa")

dev.copy(pdf, "modules/clonalDiversity_modules.pdf")
dev.off()



####################
# Clonal Homeostasis
####################


# Clusters
clonalHomeostasis(tcr.clusters, 
                  cloneType = homeostasis.var,
                  cloneCall = "aa") + 
  theme(axis.text.x = element_text(angle = 90))

dev.copy(pdf, "clusters/clonalHomeostasis_clusters.pdf")
dev.off()

# Modules
clonalHomeostasis(tcr.modules, 
                  cloneType = homeostasis.var,
                  cloneCall = "aa") + 
  theme(axis.text.x = element_text(angle = 90))

dev.copy(pdf, "modules/clonalHomeostasis_modules.pdf")
dev.off()


####################
# Clonal Proportion
####################

# Clusters
clonalProportion(tcr.clusters, 
                 split = split.var,
                 cloneCall = "aa") + 
  theme(axis.text.x = element_text(angle = 90))

dev.copy(pdf, "clusters/clonalproportion_clusters.pdf")
dev.off()


# Modules
clonalProportion(tcr.modules, 
                 split = split.var,
                 cloneCall = "aa") + 
  theme(axis.text.x = element_text(angle = 90))

dev.copy(pdf, "modules/clonalproportion_modules.pdf")
dev.off()



####################
# Clonal Overlap
####################

# Clusters
clonalOverlap(tcr.clusters, 
              cloneCall = "aa", 
              method = "overlap") + 
  theme(axis.text.x = element_text(angle = 90))

dev.copy(pdf, "clusters/clonaloverlap_clusters.pdf")
dev.off()

# Modules
clonalOverlap(tcr.modules, 
              cloneCall = "aa", 
              method = "overlap") + 
  theme(axis.text.x = element_text(angle = 90))

dev.copy(pdf, "modules/clonaloverlap_modules.pdf")
dev.off()





###################################
# Split dataset by condition
###################################

Idents(seurat.tcr) <- seurat.tcr@meta.data$condition

# US
US.seurat.tcr <- subset(seurat.tcr, ident = "US")
tcr.clusters.US <- expression2List.mod(US.seurat.tcr, group = "seurat_clusters")

# Stimulated
Stim.seurat.tcr <- subset(seurat.tcr, ident = "Stim")
tcr.clusters.Stim <- expression2List.mod(Stim.seurat.tcr, group = "seurat_clusters")




p1 <- clonalOverlap(tcr.clusters.US, 
                    cloneCall = "aa", 
                    method = "overlap") + 
  theme(axis.text.x = element_text(angle = 90))

p2 <- clonalOverlap(tcr.clusters.Stim, 
                    cloneCall = "aa", 
                    method = "overlap") + 
  theme(axis.text.x = element_text(angle = 90))


p1|p2



rm(US.seurat.tcr, tcr.clusters.US, Stim.seurat.tcr, tcr.clusters.Stim, tcr.modules, tcr.clusters)


# reset to working dir
setwd(working.dir)
```

### Clonotype frequency on UMAP
```{r Clonotype_freq}

# Ensure dir back in working
setwd(working.dir)

# Create output directory
if(!dir.exists("output/figures/TCRseq/Binned_clonotype_expansion_plots")){
  dir.create("output/figures/TCRseq/Binned_clonotype_expansion_plots",
             recursive = T)}

output.dir <- "output/figures/TCRseq/Binned_clonotype_expansion_plots"
setwd(output.dir)

###################################
# Plot dataset on UMAP projection
###################################

Idents(seurat.tcr) <- seurat.tcr@meta.data$seurat_clusters

clonetype.col <- c("Red", "Black", "blue", "orange", "lightblue")

# Vis clonotype size on UMAP
UMAPPlot(seurat.tcr, 
         group.by = "cloneType", 
         pt.size = 1) +
  scale_color_manual(values = clonetype.col, 
                     na.value="grey")


dev.copy(pdf, "Clonotype_size_UMAP.pdf")
dev.off()

# Split by condition
UMAPPlot(seurat.tcr, 
         group.by = "cloneType", 
         split.by = "condition", 
         pt.size = 1) +
  scale_color_manual(values = clonetype.col, 
                     na.value="grey")

dev.copy(pdf, "Clonotype_size_UMAP_split_condition.pdf")
dev.off()


# Split by cloneType visualise clusters
UMAPPlot(seurat.tcr, 
         group.by = "seurat_clusters", 
         split.by = "cloneType", 
         pt.size = 1, 
         cols = clust.cols) 

dev.copy(pdf, "Clonotype_size_UMAP_split_cloneType_Clusters.pdf")
dev.off()

# Split by cloneType visualise clusters
UMAPPlot(seurat.tcr, 
         group.by = "Module", 
         split.by = "cloneType", 
         pt.size = 1,
         cols = module.cols) 

dev.copy(pdf, "Clonotype_size_UMAP_split_cloneType_Modules.pdf")
dev.off()

# Ensure dir back in working
setwd(working.dir)
```

## TCR analysis - cluster-specific clonotypes {.tabset}

### Get clonotypes function
```{r Get_clonotypes_function}

get.clonotypes <- function(input.seurat, identity.column, ident.id){
  
  # identitiy column = cluster or module 
  # ident.id = a specific module or cluster ID
  
  # This function identifies all clonotypes in a specified cluster/module subset and then pulls out all occurances of these clonotypes
  
  
  
  # set idents
  Idents(input.seurat) <- input.seurat@meta.data[ , paste0(identity.column)]
  
  # Subset by ident.id
  Ident.specific.seurat <- subset(input.seurat, idents = paste0(ident.id))
  
  # get list of clonotypes in selected cluster/module
  Ident.specific.clones <- sort(table(Ident.specific.seurat@meta.data$CTaa), decreasing = TRUE)
  
  # Find all cells that have these clonotypes
  keep.logic <- input.seurat@meta.data$CTaa %in% names(Ident.specific.clones)
  cells.with.ident.sp.clones <- rownames(input.seurat@meta.data)[keep.logic]
  
  # Subset seurat by these cell IDs
  seurat.ident.specific.clones <- subset(input.seurat, cells = cells.with.ident.sp.clones)
  
  
  return(seurat.ident.specific.clones)
  
}

```

## Reduce dataset to clonotypes per cluster
```{r Cluster_clonotypes}

# set working dir
setwd(working.dir)

# Create output directories
if(!dir.exists("output/figures/TCRseq/Cluster_clonotypes")){
  dir.create("output/figures/TCRseq/Cluster_clonotypes",
             recursive = T)}

output.dir <- "output/figures/TCRseq/Cluster_clonotypes"
setwd(output.dir)

##########################################################
# Keep only one occurrence of a clonotype in each cluster
##########################################################

uniq.clonotypes <- seurat.tcr@meta.data %>% 
  dplyr::distinct(seurat_clusters, CTaa) %>%
  rownames()

clust.clonotypes.seurat <- subset(seurat.tcr, cells = uniq.clonotypes)
#3773 clonotypes 

# Set Idents
Idents(clust.clonotypes.seurat) <- clust.clonotypes.seurat@meta.data$seurat_clusters

# Cluster Clonotypes
UMAPPlot(clust.clonotypes.seurat, 
         pt.size = 1, 
         group.by = "seurat_clusters", 
         cols = clust.cols) + 
  ggtitle("Pts = Cluster clonotypes")

dev.copy(pdf, "Cluster_clonotypes.pdf")
dev.off()

# Cluster Clonotypes // No legend
UMAPPlot(clust.clonotypes.seurat, 
         pt.size = 1, 
         group.by = "seurat_clusters", 
         label = TRUE,
         label.size = 6,
         cols = clust.cols) + 
  ggtitle("Pts = Cluster clonotypes") + 
  NoLegend()

dev.copy(pdf, "Cluster_clonotypes_NoLegend.pdf")
dev.off()


# Plot number of clonotypes per cluster
x <- table(clust.clonotypes.seurat@meta.data$seurat_clusters)

barplot(x, 
        las = 2, 
        ylim = c(0, 1000), 
        ylab = "# of clonotypes",
        main = "# of clonotypes per cluster")

dev.copy(pdf, "Barplot_Num_Cluster_clonotypes.pdf")
dev.off()



############################################
# Scale pt size with clonotype frequency
############################################

Total.clonotype.scale <- clust.clonotypes.seurat@meta.data$Total_clonotype_scaled 

# Total dataset
UMAPPlot(clust.clonotypes.seurat, 
         pt.size = Total.clonotype.scale, 
         group.by = "seurat_clusters", 
         cols = clust.cols) + 
    ggtitle("Clonotypes - scaled by total occurrence")

dev.copy(pdf, "Cluster_clonotypes_scaled_by_occurrence.pdf")
dev.off()



# Total dataset // No legend
UMAPPlot(clust.clonotypes.seurat, 
         pt.size = Total.clonotype.scale, 
         group.by = "seurat_clusters", 
         label = TRUE, 
         label.size = 4,
         cols = clust.cols) + 
      ggtitle("Clonotypes - scaled by total occurrence") +
  NoLegend()

dev.copy(pdf, "Cluster_clonotypes_scaled_by_occurrence_Nolegend.pdf")
dev.off()


# Remove seurat
rm(clust.clonotypes.seurat)


# set working dir
setwd(working.dir)

```
 
## Reduce dataset to clonotypes per cluster per condition
```{r Cluster_condition_clonotypes}

# set working dir
setwd(working.dir)

# Create output directories
if(!dir.exists("output/figures/TCRseq/Cluster_condition_clonotypes")){
  dir.create("output/figures/TCRseq/Cluster_condition_clonotypes",
             recursive = T)}

output.dir <- "output/figures/TCRseq/Cluster_condition_clonotypes"
setwd(output.dir)


############################################################################
# Keep only one occurrence of a clonotype in each cluster per condition
############################################################################

uniq.clonotypes <- seurat.tcr@meta.data %>% 
  dplyr::distinct(seurat_clusters, CTaa, condition) %>%
  rownames()

clust.cond.clonotypes.seurat <- subset(seurat.tcr, cells = uniq.clonotypes)
# 4093 clonotypes

# Set Idents
Idents(clust.cond.clonotypes.seurat) <- clust.cond.clonotypes.seurat@meta.data$seurat_clusters


# Cluster & Condition clonotypes
UMAPPlot(clust.cond.clonotypes.seurat, 
         pt.size = 1, 
         group.by = "seurat_clusters", 
         split.by = "condition",
         cols = clust.cols) + 
  ggtitle("Pts = Cluster and condition clonotypes")

dev.copy(pdf, "Cluster_condition_clonotypes.pdf")
dev.off()

# Split by condition // no legend
UMAPPlot(clust.cond.clonotypes.seurat, 
         pt.size = 1, 
         group.by = "seurat_clusters", 
         split.by = "condition",
         label = TRUE, 
         label.size = 4,
         cols = clust.cols) + 
  ggtitle("Pts = Cluster and condition clonotypes") + 
  NoLegend()

dev.copy(pdf, "Cluster_condition_clonotypes_Nolegend.pdf")
dev.off()


# Barplot of # of unique clonotypes per clust per condition

x <- table(clust.cond.clonotypes.seurat@meta.data$seurat_clusters, clust.cond.clonotypes.seurat@meta.data$condition)


barplot(t(x),
        las = 2,
        beside = TRUE,
        col = Condition.cols,
        legend = TRUE,
        ylim = c(0, 600),
        ylab = "# of clonotypes", 
        main = "# of unique clonotypes per cluster per condition")

dev.copy(pdf, "Barplot_Num_Cluster_condition_clonotypes.pdf")
dev.off()


############################################
# Scale pt size with clonotype frequency
############################################

Condition.clonotype.scale <- clust.cond.clonotypes.seurat@meta.data$Clone_Clust_conditionwise_scaled 

# Split by condition
UMAPPlot(clust.cond.clonotypes.seurat, 
         pt.size = Condition.clonotype.scale, 
         group.by = "seurat_clusters", 
         split.by = "condition",
         cols = clust.cols) + 
    ggtitle("Clust/Cond Clonotype - by normed condition freq")

dev.copy(pdf, "Cluster_condition_clonotypes_scaled_by_normalised_conditionwise_freq.pdf")
dev.off()


# Split by condition // No legend
UMAPPlot(clust.cond.clonotypes.seurat, 
         pt.size = Condition.clonotype.scale, 
         group.by = "seurat_clusters", 
         split.by = "condition",
         label = TRUE, 
         label.size = 4,
         cols = clust.cols) + 
    ggtitle("Clust/Cond Clonotype - by normed condition freq") + 
  NoLegend()

dev.copy(pdf, "Cluster__condition_clonotypes_scaled_by_normalised_conditionwise_freq_Nolegend.pdf")
dev.off()


# set working dir
setwd(working.dir)

```
 
### Cluster-condition-specific clonotypes
```{r Cluster_condition_specific_clonotypes}

# Create output directories
setwd(working.dir)

if(!dir.exists("output/figures/TCRseq/Cluster_condition__specific_clonotypes")){
  dir.create("output/figures/TCRseq/Cluster_condition__specific_clonotypes",
             recursive = T)}

output.dir <- "output/figures/TCRseq/Cluster_condition__specific_clonotypes"
setwd(output.dir)


# Set Idents to Cluster ID
Idents(clust.cond.clonotypes.seurat) <- clust.cond.clonotypes.seurat@meta.data$seurat_clusters



# Get x and y lims for UMAP plots
z <- UMAPPlot(clust.cond.clonotypes.seurat, 
                 pt.size = 1,
                 group.by = "seurat_clusters",
                 split.by = "condition",
                 cols = clust.cols)

# extract x and y lims
umap.x.lim <- layer_scales(z)$x$range$range
umap.y.lim <- layer_scales(z)$y$range$range


###################################
# get Cluster specific clonotypes
###################################

Cluster.vect <- unique(clust.cond.clonotypes.seurat@meta.data$condition_clust)

for(i in 1:length(Cluster.vect)){
  
  print(paste0("Performing analysis for Cluster = ", Cluster.vect[i]))
  
  
  # Extract clonotypes found in a specific clust/condition
  clonotype.seurat <- get.clonotypes(clust.cond.clonotypes.seurat, "condition_clust", Cluster.vect[i])
  
  
  ########################
  # Visualisation 
  ########################
  
  # Overall distribution of clonotypes 
  print(UMAPPlot(clonotype.seurat, 
                 pt.size = 1,
                 group.by = "seurat_clusters",
                 split.by = "condition",
                 cols = clust.cols) + 
          xlim(umap.x.lim) + ylim(umap.y.lim) + 
          ggtitle(paste0("Clonotypes found in ", Cluster.vect[i])))
  
  dev.copy(pdf, 
           paste0("Clonotypes_found_in_", Cluster.vect[i], ".pdf"))
  dev.off()
  
  
  # Scale clonotype size
  print(UMAPPlot(clonotype.seurat, 
                 pt.size = clonotype.seurat@meta.data$Clone_Clust_conditionwise_scaled,
                 group.by = "seurat_clusters", 
                 split.by = "condition",
                 cols = clust.cols) +
          xlim(umap.x.lim) + ylim(umap.y.lim) + 
          ggtitle(paste0("Clonotypes found in ", Cluster.vect[i])))
  
  dev.copy(pdf, 
           paste0("Clonotypes_found_in_", Cluster.vect[i], "_pt_scaled.pdf"))
  dev.off()
  

############
# Barplots
############
  
# Number of unique clones
n.clonotypes <- table(clonotype.seurat@meta.data$seurat_clusters)

print(barplot(n.clonotypes, 
              ylim = c(0, 1000),
              ylab = "# of clonotypes",
              main = paste0("# clonotypes found in Cluster ", Cluster.vect[i]), 
              cex.names = 0.7, 
              las = 2))   

  dev.copy(pdf, 
           paste0("Barplot_of_clonotypes_found_in_", Cluster.vect[i], ".pdf"))
  dev.off()
  
  
  # Freq of clone overlap
  n.clonotypes <- table(clonotype.seurat@meta.data$seurat_clusters, clonotype.seurat@meta.data$condition)
  total.n.clonotypes <- table(clust.cond.clonotypes.seurat@meta.data$seurat_clusters, clust.cond.clonotypes.seurat@meta.data$condition)
  
  x <- (n.clonotypes/total.n.clonotypes)*100
  
  print(barplot(t(x), 
                ylim = c(0, 100),
                ylab = "% of clonotype overlap",
                main = paste0("% of clonotypes present in ", Cluster.vect[i]), 
                cex.names = 0.9, 
                col = Condition.cols,
                legend = TRUE,
                beside = TRUE,
                las = 2))   
  
  dev.copy(pdf, paste0("Barplot_freq_of_clonotypes_found_in_", Cluster.vect[i], ".pdf"))
  dev.off()
  
  
  # Remove unneeded variable
  rm(clonotype.seurat)
}


setwd(working.dir)


  
  
```

### Venn overlap of clonotypes
```{r Unique_shared_clonotypes_Venn}

# Create output directories
setwd(working.dir)

if(!dir.exists("output/figures/TCRseq/Venn_overlap")){
  dir.create("output/figures/TCRseq/Venn_overlap",
             recursive = T)}

output.dir <- "output/figures/TCRseq/Venn_overlap"
setwd(output.dir)

# Load package
library("RVenn")

####################################
# Overlap of all clonotypes
####################################

# Get vector of clonotypes found in US cells
US.clonotypes <- seurat.tcr@meta.data %>%
  dplyr::filter(condition == "US") %>%
  dplyr::select(CTaa)

US.clonotypes <- unique(US.clonotypes$CTaa)

length(US.clonotypes) # 1,845 clonotypes

# Get vector of clonotypes found in US cells
Stim.clonotypes <- seurat.tcr@meta.data %>%
  dplyr::filter(condition == "Stim") %>%
  dplyr::select(CTaa)

Stim.clonotypes <- unique(Stim.clonotypes$CTaa)

length(Stim.clonotypes) # 1,754 clonotypes

# Venn clonotypes

clonotype.list <- list(US = US.clonotypes, 
                       Stim = Stim.clonotypes)

clonotype.list <- Venn(clonotype.list)


ggvenn(clonotype.list,
       fill = rev(Condition.cols),
       thickness = 0) + 
  ggtitle("All Clonotypes") 

dev.copy(pdf, "Venn_overlap_All_clonotypes.pdf")
dev.off()

# Extract vectors
shared.clonotypes <- overlap(clonotype.list)
unique.US <- discern(clonotype.list, "US")
unique.Stim <- discern(clonotype.list, "Stim")

####################
# Add to metadata
####################
meta.data.vec <- seurat.tcr@meta.data %>%
  dplyr::mutate(Clonotype_overlap = case_when(CTaa %in% shared.clonotypes ~ "Shared", 
                                              CTaa %in% unique.US ~ "US", 
                                              CTaa %in% unique.Stim ~ "Stim")) %>%
  dplyr::select(Clonotype_overlap)


seurat.tcr <- AddMetaData(seurat.tcr, meta.data.vec$Clonotype_overlap, "Clonotype_overlap")


###########################
# Abundant clonotypes only
###########################

condition.min <- 20

# Get vector of clonotypes found in US cells
US.clonotypes <- seurat.tcr@meta.data %>%
  dplyr::filter(condition == "US" & Clone_conditionwise_n >= condition.min) %>%
  dplyr::select(CTaa)

US.clonotypes <- unique(US.clonotypes$CTaa)

length(US.clonotypes) #

# Get vector of clonotypes found in US cells
Stim.clonotypes <- seurat.tcr@meta.data %>%
  dplyr::filter(condition == "Stim" & Clone_conditionwise_n >= condition.min) %>%
  dplyr::select(CTaa)

Stim.clonotypes <- unique(Stim.clonotypes$CTaa)

length(Stim.clonotypes) #

# Venn overlap
clonotype.list <- list(US = US.clonotypes, 
                       Stim = Stim.clonotypes)

clonotype.list <- Venn(clonotype.list)

ggvenn(clonotype.list,
       fill = rev(Condition.cols),
       thickness = 0) + 
  ggtitle("Abundant clonotypes") 

dev.copy(pdf, "Venn_overlap_abundant_clonotypes.pdf")
dev.off()


################
# Singletons
################

# Get vector of clonotypes found in US cells
US.clonotypes <- seurat.tcr@meta.data %>%
  dplyr::filter(condition == "US" & Clone_conditionwise_n == 1) %>%
  dplyr::select(CTaa)

US.clonotypes <- unique(US.clonotypes$CTaa)

length(US.clonotypes)

# Get vector of clonotypes found in US cells
Stim.clonotypes <- seurat.tcr@meta.data %>%
  dplyr::filter(condition == "Stim" & Clone_conditionwise_n == 1) %>%
  dplyr::select(CTaa)

Stim.clonotypes <- unique(Stim.clonotypes$CTaa)

length(Stim.clonotypes)

# Venn overlap
clonotype.list <- list(US = US.clonotypes, 
                       Stim = Stim.clonotypes)

clonotype.list <- Venn(clonotype.list)

ggvenn(clonotype.list,
       fill = rev(Condition.cols),
       thickness = 0) + 
  ggtitle("Singletons") 

dev.copy(pdf, "Venn_overlap_singleton_clonotypes.pdf")
dev.off()


##############################################
# Low abundance in US and abundant in Stim
##############################################

# Get vector of clonotypes found in US cells
US.clonotypes <- seurat.tcr@meta.data %>%
  dplyr::filter(condition == "US" & Clone_conditionwise_n <= 5) %>%
  dplyr::select(CTaa)

US.clonotypes <- unique(US.clonotypes$CTaa)

length(US.clonotypes)

# Get vector of clonotypes found in US cells
Stim.clonotypes <- seurat.tcr@meta.data %>%
  dplyr::filter(condition == "Stim" & Clone_conditionwise_n >= 30) %>%
  dplyr::select(CTaa)

Stim.clonotypes <- unique(Stim.clonotypes$CTaa)

length(Stim.clonotypes)

# Venn overlap 
clonotype.list <- list(US = US.clonotypes, 
                       Stim = Stim.clonotypes)

clonotype.list <- Venn(clonotype.list)

ggvenn(clonotype.list,
       fill = rev(Condition.cols),
       thickness = 0) + 
  ggtitle("US <= 5 and Stim >= 30") 

dev.copy(pdf, "Venn_overlap_low_US_high_Stim_abundance_clonotypes.pdf")
dev.off()

```
 

